\vspace{-15mm}\fontsize{24pt}{10pt}\selectfont\textbf{SmarteRun: Coding The Portable Coach}

### 2014-05-12 final/article.tex
fixed issues with apostrophes being unicode or sum shit

\noindent The pairing of physical activity and music is a natural, common occurrence in modern society.  Today, there are many smartphone and smart device applications focused on this pairing, specifically catering towards runners. Currently, these applications make no attempt to combine music and running in a novel way, instead allowing only for both activities to occur simultaneously. As the prevalence of smart devices continues to alter the technological landscape, the opportunity for the union of this pair is becoming more feasible.  We consider psychological studies that consider the rhythmic nature of human exercise and the effects of music upon this rhythm  We then synthesize a model by which previous activity can predict future run performance and capabilities. We then analyze these predictions and discuss further improvements which can be made to our models. We conclude with the application of this predictive model to music playback and the ability to cater music at speeds which accentuate runners' performance and create the running coach athletes didn'Äôt know they had.

@@

	First, psychological theories posit that the human mind only has so much attention it can utilize at once. It follows, then, that the more one splits their attention, the less they can pay attention to any one thing. In runners, this leads to a limited amount of attention they can devote to 'Äúfeeling the pain'Äù of a run if they are listening to music. Without the music, they may just focus on how hard they are working, but with music, the music drowns out the difficulty of the exercise. This idea is backed up in research as studies found a causal link between playing music and feeling a lower rate of perceived exertion. \\

@@

	Other attempts at using past exercise data to predict future performance are scarce, however there is one major example. In 2013, the Boston Marathon had to be shutdown before many participants could finish due to an act of terrorism. In order to determine run times for the unfinished runners, a group from the Harvard School of Public Health was employed to predict performance for the incomplete times. The investigation determined the best approach was to use a K-nearest-neighbor algorithm searching for similar other runs in a dataset of past run data, and then combining them to extrapolate to the rest of the run. This method far exceeded the 'Äúconstant pace'Äù assumption others had proposed, as well as several other statistical attempts such as the ANOVA method. This method worked very well as it predicted 98\% of runs to within 10 minutes. While this is a large error for normal runs, 10 minutes is small on marathon scale. For 80\% of runners, that error is cut in half. (Hammerling)\\

	Given this research, we have validated the idea of using past data to predict future run performance as well as using music to improve and influence the runner'Äôs experience. \\

@@

	To gather data for this implementation we sampled runners with an interval of once per second. Utilizing this data we can build up our 3-dimensional graph by plotting these sampled elapsed distances over the length of the activity. We note that our z-value is simply the last x-value or the total elapsed distance at the end of our activity.  For the sake of this paper we assume that these values are correctly equivalent and runners accurately predetermine their running distance. We also utilized automated test data generation which produced pseudo-random run data constrained to reasonable performance as seen in our real test data.  We determined probabilistic changes to a runner'Äôs pace through an activity based on the ways in which real runners changed their pace over the course of a run. This data allowed us to more rigorously test the accuracy and quality of our predictions as gathering a multitude of real data was impractical for the scope of our project.\\

	Given a desired run distance z, our algorithm uses a variation of a k-closest neighbor algorithm, where we choose k closest runs to our desired activity and use these to predict our new activity.  We determine a run'Äôs closeness to another run by its z-distance, utilizing the assumption that we are considering only data from the same user.  Given these neighbors, we normalize each'Äôs time to the time interval of [0,1] (minutes) such that we can directly compare them. We note that though this is a naive direct comparison between runs, we later normalize the predictive weight of a neighboring run using exponential decay to adjust for this.  We describe how we average these k closest neighbors to produce a prediction.\\

@@

	Using this predicted graph R(x,y,z) we provide tangent-line and secant-line predictions such that we provide a recommended pace for a runner such that given their current performance they achieve, exceed, or relax compared to their usual performance.  This step essentially merges our predictive data with the runner'Äôs current activity and would be the useful user-facing data provided in a retail application of our algorithms.\\

@@

Repeat this process over multiple iterations, recording both the predicted pace and actual pace for each sample. Then calculate some measure of error (such as RMSE)
\end{itemize}

@@

	This investigation provided useful results, but it was intended mainly as a starting point. The problem of workout recommendation is the source of entire professions (i.e. personal trainers). This project assumed all runs were 'Äúgood'Äù runs and thus should be taken into account for prediction. In reality, some runs are ruined by a bad cramp, an error in data collection, or a phone call mid-run and thus should not be used to describe typical performance. Additionally, we ran our K-nearest-neighbor algorithm on the current run and matched other runs together based on just the Z variable of total run length. This could be improved if we tried to find runs similar to the current one based on the total length as well as the performance so far. This would allow us to, at least to some degree, more properly predict the longer runs as well as fix another major development that this problem needs before anyone can call it solved: atrophy during extended periods of inactivity. This app currently assumes the user is constantly training, however, a very common use case is ignored. Namely, a runner will often take a few weeks or even months off because they are particularly busy or it'Äôs winter. This would lead to a mismatch where the runner is in much worse shape than our current algorithm understands and then tells the runner to essentially sprint the rest of the run. Matching by performance so-far in the run would allow the algorithm to implicitly match on conditioning level as well. With a large enough user base, this could become very powerful as thousands of runs can be combined from a multitude of different users, much in the same way Netflix powers their suggestion engine. Furthermore, these approaches could easily be applied to many more types of exercise such as swimming or weight lifting as well. \\

	With a little bit of improvement, this type of project would be very valuable to an application such as RunKeeper. Taking the expected performance and translating it into a pace that the runner should keep would be a good next step in this project. Using assumptions based on the height of the runner would be a simple way to suggest a running pace based on speed, however an entirely separate study should be done on how to use machine learning to match a specific user to the proper pace based on where they are in the run and any other factors specific to their person and their run. This could then be a powerful tool to give runners the coach they don'Äôt even know they have.

>>>>

\noindent The pairing of physical activity and music is a natural, common occurrence in modern society.  Today, there are many smartphone and smart device applications focused on this pairing, specifically catering towards runners. Currently, these applications make no attempt to combine music and running in a novel way, instead allowing only for both activities to occur simultaneously. As the prevalence of smart devices continues to alter the technological landscape, the opportunity for the union of this pair is becoming more feasible.  We consider psychological studies that consider the rhythmic nature of human exercise and the effects of music upon this rhythm  We then synthesize a model by which previous activity can predict future run performance and capabilities. We then analyze these predictions and discuss further improvements which can be made to our models. We conclude with the application of this predictive model to music playback and the ability to cater music at speeds which accentuate runners' performance and create the running coach athletes didn't know they had.

@@

	First, psychological theories posit that the human mind only has so much attention it can utilize at once. It follows, then, that the more one splits their attention, the less they can pay attention to any one thing. In runners, this leads to a limited amount of attention they can devote to "feeling the pain" of a run if they are listening to music. Without the music, they may just focus on how hard they are working, but with music, the music drowns out the difficulty of the exercise. This idea is backed up in research as studies found a causal link between playing music and feeling a lower rate of perceived exertion. \\

@@

	Other attempts at using past exercise data to predict future performance are scarce, however there is one major example. In 2013, the Boston Marathon had to be shutdown before many participants could finish due to an act of terrorism. In order to determine run times for the unfinished runners, a group from the Harvard School of Public Health was employed to predict performance for the incomplete times. The investigation determined the best approach was to use a K-nearest-neighbor algorithm searching for similar other runs in a dataset of past run data, and then combining them to extrapolate to the rest of the run. This method far exceeded the "constant pace" assumption others had proposed, as well as several other statistical attempts such as the ANOVA method. This method worked very well as it predicted 98\% of runs to within 10 minutes. While this is a large error for normal runs, 10 minutes is small on marathon scale. For 80\% of runners, that error is cut in half. (Hammerling)\\

	Given this research, we have validated the idea of using past data to predict future run performance as well as using music to improve and influence the runner's experience. \\

@@

	To gather data for this implementation we sampled runners with an interval of once per second. Utilizing this data we can build up our 3-dimensional graph by plotting these sampled elapsed distances over the length of the activity. We note that our z-value is simply the last x-value or the total elapsed distance at the end of our activity.  For the sake of this paper we assume that these values are correctly equivalent and runners accurately predetermine their running distance. We also utilized automated test data generation which produced pseudo-random run data constrained to reasonable performance as seen in our real test data.  We determined probabilistic changes to a runner's pace through an activity based on the ways in which real runners changed their pace over the course of a run. This data allowed us to more rigorously test the accuracy and quality of our predictions as gathering a multitude of real data was impractical for the scope of our project.\\

	Given a desired run distance z, our algorithm uses a variation of a k-closest neighbor algorithm, where we choose k closest runs to our desired activity and use these to predict our new activity.  We determine a run's closeness to another run by its z-distance, utilizing the assumption that we are considering only data from the same user.  Given these neighbors, we normalize each's time to the time interval of [0,1] (minutes) such that we can directly compare them. We note that though this is a naive direct comparison between runs, we later normalize the predictive weight of a neighboring run using exponential decay to adjust for this.  We describe how we average these k closest neighbors to produce a prediction.\\

@@

	Using this predicted graph R(x,y,z) we provide tangent-line and secant-line predictions such that we provide a recommended pace for a runner such that given their current performance they achieve, exceed, or relax compared to their usual performance.  This step essentially merges our predictive data with the runner's current activity and would be the useful user-facing data provided in a retail application of our algorithms.\\

@@

\item Repeat this process over multiple iterations, recording both the predicted pace and actual pace for each sample. Then calculate some measure of error (such as RMSE)
\end{itemize}

@@

	This investigation provided useful results, but it was intended mainly as a starting point. The problem of workout recommendation is the source of entire professions (i.e. personal trainers). This project assumed all runs were "good" runs and thus should be taken into account for prediction. In reality, some runs are ruined by a bad cramp, an error in data collection, or a phone call mid-run and thus should not be used to describe typical performance. Additionally, we ran our K-nearest-neighbor algorithm on the current run and matched other runs together based on just the Z variable of total run length. This could be improved if we tried to find runs similar to the current one based on the total length as well as the performance so far. This would allow us to, at least to some degree, more properly predict the longer runs as well as fix another major development that this problem needs before anyone can call it solved: atrophy during extended periods of inactivity. This app currently assumes the user is constantly training, however, a very common use case is ignored. Namely, a runner will often take a few weeks or even months off because they are particularly busy or it's winter. This would lead to a mismatch where the runner is in much worse shape than our current algorithm understands and then tells the runner to essentially sprint the rest of the run. Matching by performance so-far in the run would allow the algorithm to implicitly match on conditioning level as well. With a large enough user base, this could become very powerful as thousands of runs can be combined from a multitude of different users, much in the same way Netflix powers their suggestion engine. Furthermore, these approaches could easily be applied to many more types of exercise such as swimming or weight lifting as well. \\

	With a little bit of improvement, this type of project would be very valuable to an application such as RunKeeper. Taking the expected performance and translating it into a pace that the runner should keep would be a good next step in this project. Using assumptions based on the height of the runner would be a simple way to suggest a running pace based on speed, however an entirely separate study should be done on how to use machine learning to match a specific user to the proper pace based on where they are in the run and any other factors specific to their person and their run. This could then be a powerful tool to give runners the coach they don't even know they have.
